Evaluating computational methods for modeling off-normal operation of gas centrifuge cascades

### 2021-02-10 motivation.tex
first review pass from abstract up until 2.5.1

With the current international tensions regarding enrichment capabilities, this
work aims to measure the effectiveness of a symmetric enrichment cascade when
used outside of its designed scope and quantify the attractiveness of such ways
to build up significant quantity of \gls{HEU}.

The present work investigates the performance of an enrichment cycle when chaining
gas enrichment cascades tuned for low enriched uranium (LEU) production from natural
uranium to instead produce HEU. Literature on the subject is limited due to its
internationally and politically sensitive nature. Three behavior models have been
implemented and used to evaluate the response of an enrichment cascade when fed
with different assays than originally designed. This work also takes advantage
of the Cyclus \cite{cyclus} fuel cycle simulator's capabilities to evaluate the
assay values at equilibrium.

>>>>

With the current international tensions regarding enrichment capabilities, this
work aims to measure the effectiveness of a symmetric enrichment cascade when
used outside of its designed scope and to quantify the attractiveness %Katie: I would be VERY VERY careful using the word attractiveness.
of such ways
to build up a \gls{SQ} of \gls{HEU}, defined by the \gls{IAEA} as \(25 \ kilograms\)\cite{noauthor_iaea_2001}.

The present work investigates the performance of an enrichment cycle when chaining
gas centrifuge enrichment cascades tuned for \gls{LEU} production from natural uranium to instead produce \gls{HEU}. Literature on the subject is limited due to its
internationally and politically sensitive nature. Three behavior models have been
implemented and used to evaluate the response of an enrichment cascade when fed
with different assays than originally designed. This work also takes advantage
of the \Cyclus \cite{cyclus} fuel cycle simulator's capabilities to evaluate the
assay values at equilibrium.

### 2021-02-10 theory.tex
first review pass from abstract up until 2.5.1

As described in \cite{glaser.2008}, the separative power of a single centrifuge
can be express as an analytical solution \cite{raetz.phd} of the
differential equation for the gas centrifuge:

@@

\begin{equation}
    \max\{[1-(\frac{r_1}{r_2})^2]^2 \times [ \ln(\frac{r_2}{r_1})]^{-1}\}
\end{equation}

The ratio can further be constrained by approximating $r_2$ as being very close
to the inner centrifuge wall with radius $a$:

\begin{equation}
    (\frac{r_1}{r_2}) \approx (\frac{r_1}{a}) = \sqrt{1 - \frac{2RT}{M}(\ln x)\frac{1}{v_{a}^{2}}}
\end{equation}

Here the gas constants are molar mass, $M$, temperature $T$, universal gas constant, $R$,
and the pressure ratio, $x$ (typically $1000:1$).
This relationship is valid when $v_a > 380 \frac{\mathrm{m}}{\mathrm{s}}$.
Otherwise, the relationship can be approximated $\{\frac{r_1}{r_2} \approx 0.534 \mid v_a \leq 380 \frac{\mathrm{m}}{\mathrm{s}}\}$.
In order to decompose the ratio into each individual radius, knowledge on one
of them is required. Glaser \cite{glaser.2008} states that $r_2$ typically ranges
from $96\%$ to $99\%$ the value of the inner centrifuge wall radius $a$. The exact
behavior of $r_2$ between these two values in operational conditions is unknown,
but a linear approximation can be made. It is assumed that $r_2$ is always at
the midpoint of these two extremes (i.e. $r_2 = 0.975 a$). Then, $r_1$ can be
found by simply multiplying this value by the ratio.

With this, all parameters of the separative power equation \ref{eq:raetz} are
defined and a value can be calculated. $\delta U$ has units similar to the feed
flow, that of $[mg/s]$.

@@

It can be shown \ref{avery} that the separative power of a single centrifuge can be
derive from the first principle and expressed as a function of the feed to
product enrichment factor, $\alpha$, the cut, $\theta$, and the feed rate, $F$:

@@

The outputs of a centrifuge relative to its input can be described with ratios
of the abundance ($R = \frac{N}{1-N}$) where the feed, product, and tail are
$N, N', N''$ respectively. Enrichment factiors of $\alpha$ (feed to product),
$\beta$ (feed to tail), and $\gamma$ (tail to product) can then be defined:

\begin{subequations}
    \label{eq_alphabeta}
    \begin{equation} \label{eq_alpha_def}
        \alpha = \frac{1-N}{N}\frac{N'}{1-N'}
    \end{equation}
    \begin{equation}\label{eq_beta-def}
        \beta = \frac{1-N''}{N''}\frac{N}{1-N}
    \end{equation}
    \begin{equation}\label{eq_gamma-def}
        \gamma = \alpha.\beta
    \end{equation}
\end{subequations}

@@

A symmetric cascade is a cascade where a stage's feed, $F$, with a tail, $T$, of the next
stage and product, $P$, of the previous one. For the cascade feeding stage
and external feed, $F_{ext}$:

@@

The method used to design a symmetric ideal cascade starts with the feeding
stage. As all the enrichment factors are equal across all the cascade, the
feeding stage is used to determine all subsequent stages.
The feed assay of the feeding stage, $N_{0}$, is fixed by the external feed assay
provided as an input.

From equation \eqref{eq_raetz} and \eqref{eq_alpha_principle} it is possible to
express $\alpha$ as a function of the feed rate $F$, the separative performance
$\delta U$ (a function of $\theta$), and the cut $\theta$:

@@

\begin{equation}\label{eq_product_assay}
    N' = \frac{\alpha R}{1 + \alpha R}
\end {equation}

Then, from mass conservation, $N = \theta N' + (1-\theta)N''$ and equation
\eqref{eq_product_assay}, it is possible to express $\beta$ as a function of
only the feed abundance, $R$, the cut $\theta$, and $\alpha$:

@@

To construct an ideal stage, a cut, $\theta$, must first be computed.
This is found iteratively, searching for the optimal cut value where $\alpha = \beta$
for a given feed and centrifuge parameters. With a separative power calculated,
an $\alpha$ and $\beta$ value can be determined for an initial cut guess. This
model assumes that the ideal cut for a stage should be between 0.1 and 0.9. The
two enrichment factors are compared, and the higher or lower cut value is chosen
by which pair of factors are closer. A new cut is determined from the chosen
factors, a new separative power is computed, and new $\alpha$ and $\beta$ values
are compared. This process continues until, to some precision, the resulting
enrichment factors are approximately equal. As illustrated in Fig. \ref{fig_a_m_b},
for a given input feed assay, the ideal feeding stage typically has a $\theta$
value between 0.45 and 0.525 when $\alpha = \beta$.

@@

In a cascade, as $\alpha_{i}$ and $\beta_{i}$ remain constant, only the value of
the cut, $\theta_{i}$, changes across the different stages of a cascade.  This
algorithm assumes that the corresponding separative power $\delta U$ (not
re-computed) can be achieved with the chosen centrifuge design, tuning other
operational parameter such as the rotation speed, the counter-current flow
ratio, etc.  Once $\theta_{i}$ is determined, it is possible to compute the
product and the tail assay.

The design of the ideal symmetric cascade is performed through 2 steps. First,
the configuration and number of stages is determined, adding stages until the
product assay of the final stage is greater or equal the product targeted assay,
and similarly the tails assay is less or equal the tails desired assay.  This
determines the number of enriching and stripping stages as well as their
enrichment properties ($N_{i}$, $N'_{i}$, $N''_{i}$,$\theta_{i}$).

@@

The first one assumes that no changes are being made to the cascade, i.e $\delta
U$, $F$ and $\theta$ are fixed across all stages. The second one assumes the cut
value at each stage is retuned to maintain the ideal state of the cascade,
$\alpha$ and $\beta$ remain fixed. The last one, described in \cite{walker.2017}
assumes the tails to product enrichment factor and the cut remain constants
(eq. \eqref{eq_gamma-def}: $\gamma = \alpha\beta$). Model behaviors and
assumptions are summarized in Tab. \ref{tab:models}.

>>>>

As described by Glaser in \cite{glaser.2008}, the separative power of a single centrifuge can be express as an analytical solution \cite{raetz.phd} of the
differential equation for the gas centrifuge:

@@

\begin{equation}
    \max \left\{ \left[1-\left(\frac{r_1}{r_2}\right)^2 \right]^2 \times \left[ \ln \left(\frac{r_2}{r_1} \right) \right]^{-1} \right\}
\end{equation}

This ratio can further be constrained by approximating $r_2$ as being very close
to the inner centrifuge wall with radius $a$:

\begin{equation}
    \left(\frac{r_1}{r_2}\right) \approx \left(\frac{r_1}{a}\right) = \sqrt{1 - \frac{2RT}{M}(\ln x)\frac{1}{v_{a}^{2}}}
\end{equation}

Here the gas constants are molar mass, $M$, temperature $T$, universal gas constant, $R$,
and the pressure ratio, $x$ (typically $1000:1$) \cite{}. %needs citation?
This relationship is valid when $v_a > 380 \frac{\mathrm{m}}{\mathrm{s}}$.
Otherwise, the relationship can be approximated $\{\frac{r_1}{r_2} \approx 0.534 \mid v_a \leq 380 \frac{\mathrm{m}}{\mathrm{s}}\}$.
In order to decompose the ratio into each individual radius, knowledge on one
of them is required. Glaser \cite{glaser.2008} states that $r_2$ typically ranges
from $96\%$ to $99\%$ the value of the inner centrifuge wall radius $a$. The exact
behavior of $r_2$ between these two values in operational conditions is unknown,
but a linear approximation can be made. It is assumed that $r_2$ is always at
the midpoint of these two extremes (i.e. $r_2 = 0.975 a$). Then, $r_1$ can be
found by simply multiplying this value by the ratio.

With this, all parameters of the separative power equation \ref{eq:raetz} are
defined and a value can be calculated. Separative power, $\delta U$, has units similar to the feed flow, that of $[mg/s]$.

@@

It can be shown \cite{avery} that the separative power of a single centrifuge can be
derived from first principles and expressed as a function of the feed-to-product enrichment factor, $\alpha$, the cut, $\theta$, and the feed rate, $F$:
%I think we should replicate at least some of that here

@@

The outputs of a centrifuge relative to its input can be described with ratios
of the abundance ($R = \frac{N}{1-N}$) where the feed, product, and tail enrichment $N, N', N''$ respectively. Enrichment factors of $\alpha$ (feed-to-product),
$\beta$ (feed-to-tail), and $\gamma$ (tail-to-product) can then be defined:

\begin{subequations}
    \label{eq_alphabeta}
    \begin{equation} \label{eq_alpha_def}
        \alpha = \frac{1-N}{N}\frac{N'}{1-N'}
    \end{equation}
    \begin{equation}\label{eq_beta-def}
        \beta = \frac{1-N''}{N''}\frac{N}{1-N}
    \end{equation}
    \begin{equation}\label{eq_gamma-def}
        \gamma = \alpha \, \beta
    \end{equation}
\end{subequations}

@@

A symmetric cascade is a cascade where a stage is fed using the tail, $T$, of the next stage and the product, $P$, of the previous one. The the cascade feeding stage, $F_{i}$ flow, with external feed, $F_{ext}$, is given by:

@@

Designing a symmetric ideal cascade beings with the feeding
stage. As all the enrichment factors are equal across all the cascade, the
feeding stage is used to determine all subsequent stages.
The feed assay of the feeding stage, $N_{0}$, is fixed by the external feed assay
provided as an input.

From equation \eqref{eq_raetz} and \eqref{eq_alpha_principle} it is possible to
express the feed-to-product enrichment factor $\alpha$ as a function of the feed rate $F$, the separative performance
$\delta U$ (a function of $\theta$), and the cut $\theta$:

@@

\begin{equation}\label{eq_product_assay}
    N' = \frac{\alpha \frac{N}{1-N}}{1+\alpha \frac{N}{1-N}} = \frac{\alpha R}{1 + \alpha R}
\end {equation}

Then, from mass conservation, $N = \theta N' + (1-\theta)N''$ and equation
\eqref{eq_product_assay}, it is possible to express the feed-to-tail enrichment factor $\beta$ as a function of only the feed abundance, $R$, the cut $\theta$, and the feed-to-product enrichment factor, $\alpha$:

@@

To construct an ideal stage, a cut, $\theta$, must first be computed.
This is found iteratively, searching for the optimal cut value where $\alpha = \beta$
for given feed and centrifuge parameters. With a separative power calculated,
an $\alpha$ and $\beta$ value can be determined for an initial cut guess. This
model assumes that the ideal cut for a stage should be between 0.1 and 0.9. The
two enrichment factors are compared, and the higher or lower cut value is chosen
by which pair of factors are closer. A new cut is determined from the chosen
factors, a new separative power is computed, and new $\alpha$ and $\beta$ values
are compared. This process continues until, to some precision, the resulting
enrichment factors are approximately equal. As illustrated in Fig. \ref{fig_a_m_b},
for a given input feed assay, the ideal feeding stage typically has a $\theta$
value between 0.45 and 0.525 when $\alpha = \beta$.

@@

In a cascade, as the stage feed-to-product enrichment factor $\alpha_{i}$ and stage feed-to-tail enrichment factor $\beta_{i}$ remain constant, only the value of
the cut, $\theta_{i}$, changes across the different stages of a cascade.  This
algorithm assumes that the corresponding separative power $\delta U$ (not
re-computed) can be achieved with the chosen centrifuge design by tuning other
operational parameters such as the rotation speed, counter-current flow
ratio, etc.  Once $\theta_{i}$ is determined, it is possible to compute the
product and the tail assays.

The design of the ideal symmetric cascade is performed through 2 steps. First,
the configuration and number of stages is determined, adding stages with product assay $N'_i$ calculated using equation \ref{eq_product_assay}, until the
product assay of the final stage is greater than or equal to the product targeted assay. The stage tails assay $N''_i$ is calculated similarly, until it is less than or equal to the tails desired assay.  This determines the number of enriching and stripping stages as well as their enrichment properties ($N_{i}$, $N'_{i}$, $N''_{i}$,$\theta_{i}$).

@@

The first method, A, assumes that no changes are being made to the cascade, i.e $\delta U$, $F$ and $\theta$ are fixed across all stages. The second method, B, assumes the cut value $\theta$ is re-tuned at each stage to maintain the ideal state of the cascade, while $\alpha$ and $\beta$ remain fixed. The last method, C, described in \cite{walker.2017} assumes the tails-to-product enrichment factor $\gamma$ and the cut $\theta$ remain constants
(eq. \eqref{eq_gamma-def}: $\gamma = \alpha\beta$). Model behaviors and
assumptions are summarized in Tab. \ref{tab:models}.
